# Reproducible Research by JHU: Peer Assessment 1

#### Linh Dang - 5 Jan 2020  

This is my submission for the first project in "Reproducible Research" by JHU on Coursera. The data used is `activity.csv`, which is unzipped from this zip file: [activity.zip](https://github.com/linhdang0191/RepData_PeerAssessment1/blob/master/activity.zip)

For more information about the data, read [README](https://github.com/linhdang0191/RepData_PeerAssessment1/blob/master/README.md).

## Loading and Preprocessing Data
Assume that we already have `activity.csv` in the working directory, use `read.csv()` to read it. 

```{r read_data}
# Read the data
data <- read.csv('activity.csv')
```

We obtaine a dataframe containing 17,568 observations with 3 variables: `steps`,`date`, and `interval`.

```{r}
str(data)
```

Import necessary packages:
```{r message = FALSE, warning = FALSE}
library(dplyr) # for easier data manipulation
library(ggplot2) # for plotting
```

## What is mean total number of steps taken per day?

1. Calculate total number of steps taken per day

2. Make a histogram of total number of steps taken per day

3. Calculate the mean and median of total number of steps taken per day

First, let's create a new dataset with sum of steps taken per day.

```{r stepPerDay}
stepPerDay <- data %>% group_by(date) %>% summarise(steps = sum(steps))
```


We obtain a dataframe with 61 observations and 2 variables (with some NA rows).

```{r}
stepPerDay
```


Then make a histogram.

```{r plot1, message = FALSE, warning = FALSE}
g <- ggplot(stepPerDay, aes(steps)) + geom_histogram(col = 'darkred', fill = 'pink', binwidth = 1000) +  labs(title = "Histogram Of Steps Taken Per Day", x = "Steps", y = "Frequency") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks = seq(0,20000,2500)) + scale_y_continuous(breaks = seq(0,20,2))
# Show the plot
g
```


Mean and median of the total number of steps taken per day (ignore the NAs).

```{r mean_and_median1}
# Mean 
mean(stepPerDay$steps, na.rm = TRUE)

# Median
median(stepPerDay$steps, na.rm = TRUE)
```

As we can observe, the mean and median are 2 close values. 

## What is the average daily activity pattern?

1. Make a time-serie plot of 5-minute interval (x-axis) and the averaged steps taken across all days (y-axis)

2. Find the interval containing the maximum number of steps

Let's create a new dataframe aggregating number of steps taken in each interval and calculate their average values.


```{r stepPerInterval}
stepPerInterval <- data %>% group_by(interval) %>% summarise(averageStep = mean(steps,na.rm = TRUE))
```


We obtain a new dataframe with 288 observations and 2 variables.

```{r}
str(stepPerInterval)
```


Make a time-serie plot of 5-minute interval and the averaged steps taken across all days.

```{r plot2}
g1 <- ggplot(stepPerInterval, aes(interval,averageStep)) + geom_line(col = "darkred") + labs(title = "Average Step Taken Per Interval", x = "Interval", y = "Average Step") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks = seq(0,4000,500))
# Show the plot 
g1
```

Find the interval containing maximum number of steps.

```{r}
stepPerInterval[which.max(stepPerInterval$averageStep),]
```

The interval `835` contains the maximum number of steps (`206.1698113`).

## Imputing missing values

1. Calculate and report the number of missing values in the dataset

2. Devise a strategy to fill in missing values

3. Create a new dataset with missing values filled in 

4. Make a histogram with the total number of steps taken per day, calculate mean and median with the new dataset


Observing the original dataset, we could conclude that the only column with NAs is `steps`. Therefore, we could apply `is.na()` on `steps` only. 

Since `is.na()` will label rows with NA as `TRUE` and the other with `FALSE`, use `sum()` to calculate the number of NAs.


```{r num_NA}
sum(is.na(data$steps))
```


It is optional, but I would calculate the percentage of NAs in the original dataset.


```{r}
sum(is.na(data$steps))/NROW(data$steps)
```

About 13% values of `steps` are NAs. 

To impute missing values, there are multiple strategies to choose from. One common option to deal with numerical missing values is filling them with suitable means/medians. 

However, I would use a method called MICE (Multivariate Imputation by Chained Equation), which enables us to fill in NAs with more "reasonable" values. 

For more information, please reference to this [paper](https://www.jstatsoft.org/article/view/v045i03) and [R Documentation for mice](https://www.rdocumentation.org/packages/mice/versions/3.7.0/topics/mice).

To use this method, let's import `mice`

```{r message = FALSE, warning = FALSE}
library(mice)
```

Fill in NAs and create a new dataset. The process will take a while. 

```{r completeData, message = FALSE, warning = FALSE}
completeData <- complete(mice(data),1)
```

We receive a new dataset which is equal to the original dataset with missing values filled in. 

```{r}
str(completeData)
```


Now, we make a histogram of total number of steps taken per day with the new dataset.  

Create a dataframe with total number of steps taken per day.

```{r c_stepPerDay}
c_stepPerDay <- completeData %>% group_by(date) %>% summarise(steps = sum(steps))
```

Make a new histogram.

```{r plot3}
g2 <- ggplot(c_stepPerDay, aes(steps)) + geom_histogram(col = 'darkgreen', fill = 'lightgreen', binwidth = 1000) +  labs(title = "Histogram Of Steps Taken Per Day", x = "Steps", y = "Frequency") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks = seq(0,20000,2500)) + scale_y_continuous(breaks = seq(0,20,2))
# Show the plot
g2
```

By imputing missing values, we could see some changes are introduced to the histogram. To be more specific, we could see how the steps 0 - 2500 and 7500 - 10000 - 12500 has changed, which means the imputation has altered the frequency of some ranges of step in the dataset. 


Calculate the mean and median of total number of steps taken per day.

```{r mean_and_median2}
# Mean
mean(c_stepPerDay$steps)
# Median 
median(c_stepPerDay$steps)
```


The difference between mean and median in the new dataset is bigger than in the original dataset.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable with 2 levels: `weekday` and `weekend`.

2. Make a panel time-serie plot containing the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)


My approach for the factor variable is: create a vector to indicate weekdays; use that vector to create a factor based on logical value returend from the condition (if values returned by `weekdays()` is in the `weekdays` vector, then it is `weekday`, else, it is `weekend`)


```{r factor_weekday}
completeData$date <- as.Date(completeData$date)

# create a vector to indicate weekdays 
weekdays <- c('Monday', "Tuesday", "Wednesday", "Thursday", "Friday")

# create a column indicating day type 
completeData$dayType <- factor(weekdays(completeData$date) %in% weekdays, levels = c(TRUE,FALSE), labels = c("weekday", "weekend"))

# I create one more new column to indicate the day name
completeData$day <- factor(weekdays(completeData$date))
```

We now check if the factor variables have been created.

```{r}
str(completeData)
```

To make the panel containing time-serie plots, first we create a new dataframe of 5-minute interval with averaged steps taken across weekdays and weekends. 

```{r day_stepPerInterval}
day_stepPerInterval <- completeData %>% group_by(interval,dayType) %>% summarise(averageStep = mean(steps))
```

The result is a datafram containg 576 observations and 3 variables. 

``` {r}
str(day_stepPerInterval)
```

Make the plot.

```{r plot4}
g3 <- ggplot(day_stepPerInterval,aes(interval,averageStep)) + geom_line(col = "darkblue") + facet_grid(dayType~.) + labs(title = "Average Step Taken Per Interval", x = "Interval", y = "Average Step") + theme(plot.title = element_text(hjust = 0.5))
# Show the plot
g3
```

Okay, done!

